Protecting causal effects

### 2016-05-08 privacy/causal_differential_privacy.tex
adds some formalisms

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}

@@

Assume we add noise to the dataset via a single point crossover process (see paper). The goal is to prevent reliable estimation of causal effects without effecting our ability to predict a particular target variable. Some specific questions:

\begin{itemize}
\item Under what circumstances (graph structures) it is possible to disrupt inference of a particular causal effect
\item How can we maxiumully disrupt this inference (ie is there a cut that adds more noise for a given amount of shuffling of the data)
\item What about if we want to disrupt a specific set of causal inference questions
\item What about if we want to disrupt as many causal queries as possible.
\end{itemize}

The single point crossover process divides the variables into two groups and makes estimates of the form $P(X_1|X_2)$ unreliable if $X_1$ and $X_2$ are in different partitions. We can think of the process as drawing a cut through the graph.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}

\usetikzlibrary{arrows,positioning,backgrounds} 
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\tiny,
           text centered,
           scale=.8,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\pt}{\mathcal P}

@@

Assume we add noise to the dataset via a single point crossover process (see paper). The goal is to prevent reliable estimation of causal effects without effecting our ability to predict a particular target variable. Some example questions:

\begin{itemize}
\item Under what circumstances (graph structures) it is possible to disrupt inference of a particular causal effect \textcolor{red}{always provided there is a direct causal relationship between the exposure variable $X$ and outcome variable $Y$, ie $P(Y|do(X))\neq P(Y)$} 
\item How can we maxiumully disrupt this inference (ie is there a cut that adds more noise for a given amount of shuffling of the data) \textcolor{red}{data dependent}
\item What about if we want to disrupt a specific set of causal inference questions
\item What about if we want to disrupt as many causal queries as possible.
\end{itemize}

\section{Problem Statement and Definitions}

\begin{itemize}
\item Let $G$ be a causal directed acyclic graph over vertices $V$ and edges $E$.
\item Let $S = \set{(X_1,Y_1),(X_2,Y_2),...,(X_m,Y_m): X_i,Y_i \in V,X_i \neq Y_i}$ be a set of pairs of vertices in $G$.
\item Each pair of vertices $(X_i,Y_i)$ in $S$ represents the causal query $P(Y_i|do(X_i))$
\item A set of variables $Z \subset V$ is an adjustment for a query $(X,Y)$ if $X, Y \notin Z$ and $P(Y|do(X)) = \sum_Z P(Y|X,Z)P(Z)$
\item An adjustment for a query is minimal if it does not contain another adjustment as a proper subset.
\item The Single Point Crossover Process partitions the variables $V$ into two disjoint sets, $\pt_1$ and $\pt_2$ such that $\pt_1 \cup \pt_2 = V$. Let $\pt_X$ denote the partition containing the variable $X$.
\end{itemize}

\subsection{Discouraging inference of a set of causal queries via adjustment}

\begin{lemma}
One of the following holds.
\begin{enumerate}
\item $P(Y|do(X)) = P(Y)$
\item There is no set of variables that satisfy the back-door criterion for identifying $P(Y|do(X))$ 
\item There is at least one set of variables $Z$ that are a minimal adjustment for $P(Y|do(X))$, $Z$ may be the empty set. 
\end{enumerate}
\end{lemma}

For the remainder we assume that all queries we wish to jam are of case 3, because in case 1, there is no causal effect of $X$ on $Y$ to discover and in case 2, the causal effect of $X$ on $Y$ can already not be inferred from the data and no way in which we partition the data will render it identifiable.

\begin{lemma}Let $Z$ be a set of variables that is a minimal adjustment for the causal query $(X,Y)$. Inferring $P(Y|do(X))$ via adjusting for $Z$ is jammed if $Y \cup Z \not\subset \pt_X$. In other words, if at least one of the variables in $X \cup Y \cup Z$ is in a different a different partition to the others.

\color{red} This stems from the fact that the expression for calculating $P(Y|do(X))$ via adjusting for $Z$ contains the factor $P(Y|X,Z)$.  I'm assuming the single point crossover process gives us this result. We need to a clear definition of 'jammed' in terms of the crossover-process.
\end{lemma}

\begin{lemma} Let $\boldsymbol{Z}$ be the set of minimal adjustments for $(X,Y)$. The causal query $(X,Y)$ is jammed if inferring $P(Y|do(X))$ via $Z$ is jammed for $\forall Z \in \boldsymbol{Z}$ 
\end{lemma}

\subsubsection{An algorithm for finding a cut that renders jams a set of causal queries $S$}
\begin{enumerate}
\item For each query $(X_i,Y_i)\in S$, find the set of minimal adjustments $\boldsymbol{Z}_i$  
\item Construct the set of sets $Q = \set{X_i \cup Y_i \cup Z_{ik} : 1 \leq i \leq m, 0 \leq k \leq |\boldsymbol{Z}_i|}$
\item Find a partition such that $(Q_i \nsubseteq \pt_1)\land (Q_i \nsubseteq \pt_2)\qquad \forall Q_i \in Q$
\end{enumerate}

For example in figure \ref{fig:minimal_sets_example}, with $S = \set{(A,B),(E,B),(C,B),(E,A)}$

\eq{
&S_1 = (A,B),\, \boldsymbol{Z}_1 = \set{\set{C, E},\set{D,E} \set{D,F}} \\
&S_2 = (E,B),\, \boldsymbol{Z}_2 = \set{\set{C,D}} \\
&S_3 = (C,B),\, \boldsymbol{Z}_3 = \set{} \\
&S_4 = (E,A),\, \boldsymbol{Z}_4 = \set{C} \\
&Q = \set{\set{A,B,C,E},\set{A,B,D,E},\set{A,B,D,F},\set{E,B,C,D},\set{C,B},\set{E,A,C}}
}

We could jam $S$ with any partition of the form $\set{A,B,...},\set{C,E,F,...}$.$S_2,S_3$ and $S_4$ are jammed because their $X$ and $Y$ variables are in different partitions. $S_1$ is jammed because, although $A$ and $B$ are in the same partition, for each of the three minimal adjustment sets, at least one variable is not in $\pt_A$. $D$ could be placed in either partition. This is not the only solution, $\set{A,C,...},\set{B,E,...}$ is another. Note than in any solution $B$ and $C$ must be in different partitions as this is the only way to jam $S_3$.

The number of minimal adjustments for a given query can grow exponentially with the number of nodes in the graph. There is an algorithm to enumerate them that requires $O(n^3)$ per adjustment \cite{textor2012adjustment} .

\begin{figure}
\caption{}
\label{fig:minimal_sets_example}
\centering
   \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
%every node/.style={scale=0.6}
 %nodes
\node[main node](1){C};
\node[main node, below right=of 1](3){E};
\node[main node, above right=of 3](2){D};
\node[main node, below right=of 3](4){F};
\node[main node, below right=of 4](5){B};
\node[hidden, below left=of 1](7){};
\node[main node, below left=of 7](6){A};
 \path[every node/.style={font=\tiny}]
    (1) edge (2) edge (3) edge (6)
    (6) edge (5)
    (2) edge (3) edge (5)
    (3) edge (4) edge (6)
    (4) edge (5);
\end{tikzpicture}
\end{figure}

\subsubsection{Some things we can say}

\begin{itemize}
\item If $\boldsymbol{X} = \set{X_1 ... X_m}$ and $\boldsymbol{Y} = \set{Y_1...Y_m}$ are disjoint sets, then any partition with $\boldsymbol{X} \subseteq \pt_1$ and $\boldsymbol{Y} \subseteq \pt_2$ will jam $S$. This holds trivially if $|S|$ = 1.
\item The hardest case to jam is when $\boldsymbol{Z}_i = {} \forall i$
\item If $|S| = 2$ we can always jam $S$ as either $\boldsymbol{X}$ and $\boldsymbol{Y}$ will be disjoint or a variable $A$ will appear in both $S_1$ and $S_2$, in which case we can let $\pt_1 = \set{A}$.
\end{itemize}

\bibliographystyle{apalike}
\bibliography{/home/finn/phd/Causality-Identify}
